Cataclysmic variables are a key population of gravitational wave sources for LISA

### 2023-06-16 src/tex/ms.tex
adding in changes from Simo

% Abstract of the paper
\begin{abstract}
AM~CVn stars are well established as an important population of gravitational wave (GW) sources for the planned \lisa\ mission. The less compact – but more numerous – cataclysmic variables (CVs)\textbf{, binaries containining white dwarfs which are observed to accrete hydrogen-rich material from a companion,} have not been studied carefully in this context. Here, we estimate the predicted GW signals from the Galactic population of cataclysmic variables and evaluate their significance for \lisa. First, we find that at least three known systems are expected to produce strong enough signals to be individually resolved within the first four years of \lisa's operation. Second, cataclysmic variables will contribute significantly to the \lisa\ Galactic binary background, thus limiting the mission's sensitivity in the relevant frequency band. Third, we predict a spike in the unresolved GW background at a frequency corresponding to the cataclysmic variable minimum orbital period. This excess noise may impact the detection of other low signal-to-noise systems near this characteristic frequency. Fourth, we note that the amplitude and location of the excess noise spike associated with \textbf{$P_{\rm{min}}\sim80~\mathrm{min}$} can be used to measure the cataclysmic variable space density and period bounce location with complementary and simple GW biases compared to the biases and selection effects that plague samples selected from electromagnetic signals. Our results highlight the need to explicitly include the Galactic cataclysmic variable population in \lisa\ mission planning, both as individual GW sources and generators of background noise, as well as the exciting prospect of characterising the cataclysmic variable population through their GW emission.
\end{abstract}

@@

%The period minimum is also associated with a change in the structure of the %donor star. 
Below the period gap, the thermal time-scale of the donor increases significantly faster than the mass-loss time-scale. \textbf{This causes the donor's mass-radius index, $\mathrm{d}\ln R/\mathrm{d}\ln M$, to evolve from a near-thermal equilibrium value of $\zeta_{th} \simeq 1$ towards the adiabatic value of $\zeta_{ad} = -1/3$, where mass loss occurs at faster rate than the radius can respond}. However, the geometry of a semi-detached binary system implies that the donor must also satisfy an orbital period - density relation. It is easy to show from this that the orbital period must start to increase again once the mass-radius index of the donor reaches $\zeta = 1/3$ \citep[see e.g.][]{knigge11}. In practice, this happens at $P_{\rm{min}} \simeq 80$~min and roughly coincides with the transition of the donor from a stellar to a sub-stellar object \textbf{which is no longer fusing in its core}. The period minimum is expected -- and observed -- to coincide with a pile-up in the CV period distribution, the so-called "period spike" \citep[][]{gansicke09}.

@@

\textbf{The first volume-limited of CVs was selected using parallax measurements provided by \gaia was released by \cite{pala20}.} The sample consists of $42$ CVs within $150\,\rm{pc}$ and is estimated by them to be $\approx 75\%$ complete. Fig.~\ref{fig:porb} shows the orbital period distribution for these 42 systems in both cumulative (dotted orange curve) and differential forms (thick orange curve).

@@

%\kb{commented out what I think is a rogue \pjg}
In order to determine how much the $1~\rm{kpc}$ CV population contributes to the Galactic GW foreground for \lisa\, we must include the effects of the Galactic double WD population as well as the \lisa\ constellation orbit. We do this using \texttt{ldasoft} \citep{ldasoft}. For this study, we use the same Galactic population as was used for the LDC2a-v1 data set provided by the \lisa\ Data Challenges (LDCs)~\citep{le_jeune_maude_2022_7132178} which \textbf{consists of both detached and interacting} double WD binaries.  We consider two population cases, one which contains \textbf{only the double WD binaries from the LDC} and one which adds our $1~\rm{kpc}$ mock population of CVs before modeling the \lisa\ response to the combined signal of each Galactic binary after four years of operations. Together, these two simulations allow us to capture the effects of considering the $1~\rm{kpc}$ CV population in the Galactic foreground.

@@

It is also interesting to note that there may be the possibility of detecting other CV systems through their GW emission which have been missed due to electromagnetic detection biases and not included in the current $150~\rm{pc}$ volume-limited sample. In our simulated $150~\rm{pc}$ population, which represents a single statistical realization, we find $12$ CVs in addition to the $42$ systems included in the \citet{pala20} catalog. Of those $12$, the CV with the largest ASD has a distance of $65~\rm{pc}$ and a GW frequency $f_{\rm{GW}} = 0.26~\rm{mHz}$. This system is likely an outlier, but is consistent with our adopted space density and orbital period distribution. Perhaps more important are the CVs with distances beyond $100~\rm{pc}$ that have orbital periods near $P_{\rm{min}}$, thus aiding in their GW detection while having dimmer EM signatures due to the low mass of the donor star. This population could aid in providing tighter constraints on the local space density of the CV population. Finally, we note that it is almost certainly impossible for LISA to observe any orbital evolution for any of the individually resolved CVs based on the models of \citet{knigge11}. \textbf{A lack of an observed orbital evolution will also limit a distance measurement or precise sky position.} This is in stark contrast to AM CVn systems \citep[e.g.][]{Nelemans2004, Kremer2017, Breivik2018} and suggests that the bulk of the scientific return for the CV population will come from observations of its contribution to the LISA confusion foreground.

@@

Given the high space density of CVs within $1~\rm{kpc}$ and lack of potential to observe the orbital evolution of any individual source, it is instructive to evaluate the impact of this population of the \lisa\ confusion noise. 
Fig.~\ref{fig:isd} shows the GW noise profile and confusion obtained from \texttt{ldasoft} \citep{ldasoft} for the two simulations of the Galactic population of detached DWDs discussed in Section~\ref{sec:mock_CV}. Since \texttt{ldasoft} analyzes each of the time-delay interferometry channels in the LISA data stream, we consider the ASD from the A channel only. Specifically, Fig.~\ref{fig:isd} shows the ASD from the A channel of the simulated data (dark gray), the residual after all resolvable binaries have been removed from the data (light gray), the confusion noise level from the LDC population (orange) and the excess confusion noise from the CV population (green). The difference between the blue and the orange curve is the contribution to the \lisa\ foreground signal due to unresolved CVs. Interestingly we note that there are $\mathcal{O}(10^2)$ resolvable CVs in this mock population, as indicated by the difference between the dark and light gray traces at frequencies where the CV noise dominates.

>>>>

% Abstract of the paper
\begin{abstract}
AM~CVn stars are well established as an important population of gravitational wave (GW) sources for the planned \lisa\ mission. The less compact – but more numerous – cataclysmic variables (CVs)\textbf{, binaries containining white dwarfs which are observed to accrete hydrogen-rich material from a companion,} have not been studied carefully in this context. Here, we estimate the predicted GW signals from the Galactic population of CVs and evaluate their significance for \lisa. First, we find that at least three known systems are expected to produce strong enough signals to be individually resolved within the first four years of \lisa's operation. Second, CVs will contribute significantly to the \lisa\ Galactic binary background, thus limiting the mission's sensitivity in the relevant frequency band. Third, we predict a spike in the unresolved GW background at a frequency corresponding to the CV minimum orbital period. This excess noise may impact the detection of other low signal-to-noise systems near this characteristic frequency. Fourth, we note that the amplitude and location of the excess noise spike associated with \textbf{$P_{\rm{min}}\sim80~\mathrm{min}$} can be used to measure the CV space density and period bounce location with complementary and simple GW biases compared to the biases and selection effects that plague samples selected from electromagnetic signals. Our results highlight the need to explicitly include the Galactic CV population in the \lisa\ mission planning, both as individual GW sources and generators of background noise, as well as the exciting prospect of characterising the CV population through their GW emission.
\end{abstract}

@@

%The period minimum is also associated with a change in the structure of the %donor star. 
Below the period gap, the thermal time-scale of the donor increases significantly faster than the mass-loss time-scale. \textbf{This causes the donor's mass-radius index, $\mathrm{d}\ln R/\mathrm{d}\ln M$, to evolve from a near-thermal equilibrium value of $\zeta_{th} \simeq 1$ towards the adiabatic value of $\zeta_{ad} = -1/3$, where mass loss occurs at a faster rate than the radius can respond}. However, the geometry of a semi-detached binary system implies that the donor must also satisfy an orbital period - density relation. It is easy to show from this that the orbital period must start to increase again once the mass-radius index of the donor reaches $\zeta = 1/3$ \citep[see e.g.][]{knigge11}. In practice, this happens at $P_{\rm{min}} \simeq 80$~min and roughly coincides with the transition of the donor from a stellar to a sub-stellar object \textbf{which is no longer fusing in its core}. The period minimum is expected -- and observed -- to coincide with a pile-up in the CV period distribution, the so-called "period spike" \citep[][]{gansicke09}.

@@

\textbf{The first volume-limited of CVs was selected using parallax measurements provided by \gaia\ was released by \cite{pala20}.} The sample consists of $42$ CVs within $150\,\rm{pc}$ and is estimated by them to be $\approx 75\%$ complete. Fig.~\ref{fig:porb} shows the orbital period distribution for these 42 systems in both cumulative (dotted orange curve) and differential forms (thick orange curve).

@@

%\kb{commented out what I think is a rogue \pjg}
In order to determine how much the $1~\rm{kpc}$ CV population contributes to the Galactic GW foreground for \lisa\, we must include the effects of the Galactic detached double WD population as well as the \lisa\ constellation orbit. We do this using \texttt{ldasoft} \citep{ldasoft}. For this study, we use the same Galactic population as was used for the LDC2a-v1 data set provided by the \lisa\ Data Challenges (LDCs)~\citep{le_jeune_maude_2022_7132178} which \textbf{consists of both detached and interacting} double WD binaries. We consider two population cases, one which contains \textbf{only the double WD binaries from the LDC} and one which adds our $1~\rm{kpc}$ mock population of CVs before modeling the \lisa\ response to the combined signal of each Galactic binary after four years of operations. Together, these two simulations allow us to capture the effects of considering the $1~\rm{kpc}$ CV population in the Galactic foreground.

@@

It is also interesting to note that there may be the possibility of detecting other CV systems through their GW emission which have been missed due to electromagnetic detection biases and not included in the current $150~\rm{pc}$ volume-limited sample. In our simulated $150~\rm{pc}$ population, which represents a single statistical realisation, we find $12$ CVs in addition to the $42$ systems included in the \citet{pala20} catalog. Of those $12$, the CV with the largest ASD has a distance of $65~\rm{pc}$ and a GW frequency $f_{\rm{GW}} = 0.26~\rm{mHz}$. This system is likely an outlier, but is consistent with our adopted space density and orbital period distribution. Perhaps more important are the CVs with distances beyond $100~\rm{pc}$ that have orbital periods near $P_{\rm{min}}$, thus aiding in their GW detection while having dimmer EM signatures due to the low mass of the donor star. This population could aid in providing tighter constraints on the local space density of the CV population. Finally, we note that it is almost certainly impossible for \lisa\ to observe any orbital evolution for any of the individually resolved CVs based on the models of \citet{knigge11}. \textbf{A lack of an observed orbital evolution will also limit a distance measurement or precise sky position.} This is in stark contrast to AM CVn systems \citep[e.g.][]{Nelemans2004, Kremer2017, Breivik2018} and suggests that the bulk of the scientific return for the CV population will come from observations of its contribution to the \lisa\ confusion foreground.

@@

Given the high space density of CVs within $1~\rm{kpc}$ and lack of potential to observe the orbital evolution of any individual source, it is instructive to evaluate the impact of this population of the \lisa\ confusion noise. 
Fig.~\ref{fig:isd} shows the GW noise profile and confusion obtained from \texttt{ldasoft} \citep{ldasoft} for the two simulations of the Galactic population of detached DWDs discussed in Section~\ref{sec:mock_CV}. Since \texttt{ldasoft} analyzes each of the time-delay interferometry channels in the \lisa\ data stream, we consider the ASD from the A channel only. Specifically, Fig.~\ref{fig:isd} shows the ASD from the A channel of the simulated data (dark gray), the residual after all resolvable binaries have been removed from the data (light gray), the confusion noise level from the LDC population (orange) and the excess confusion noise from the CV population (green). The difference between the blue and the orange curve is the contribution to the \lisa\ foreground signal due to unresolved CVs. Interestingly we note that there are $\mathcal{O}(10^2)$ resolvable CVs in this mock population, as indicated by the difference between the dark and light gray traces at frequencies where the CV noise dominates.
